Leaf traits drive differences in biomass partitioning among major plant functional types

### 2015-08-13 manuscript.Rnw
Figure refs fixed

Here, we use a unique, new database \citep{falster2015} (Fig. 1) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms. A second objective was to study relationships between biomass partitioning and climate variables or biome at a global scale. The dataset includes observations of biomass and size metrics for individual plants, compiled from 175 studies across nine vegetation types (Fig. S1), across the three major biomes (boreal, temperate and tropical). In this paper we focus on field-grown woody plants, spanning the entire size range of woody plants (0.01 - \textgreater 100 m height). One key challenge is thus to account for the very large size variation commonly found in any allometric variable \citep{niklas1994}. We do this by fitting a statistical model that does no assume a functional dependence on plant size, allowing us to study PFT patterns at a common plant height.

@@

Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig. 2a and Fig. S2). When corrected for plant height by estimating LMF at a common plant height, we found that LMF was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig. 2b). As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig. 2c and Fig. S3). Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig. 2d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig. 2e).

@@

We found that LMF differs predictably among PFTs, however there was considerable variation in all studied variables between species within PFTs (Fig. S5). It is thus pertinent to understand whether we can improve on a PFT-based classification by including other traits that affect biomass partitioning. To do this, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq. \ref{eq:M2}).

@@

The ratio $M_F/A_S$ is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}, using records of stem diameter commonly recorded on long-term monitoring plots. The correlation between $M_F/A_S$ and LMA was found to hold also within PFTs (Fig. 3a), explaining 30\% of the variation in $M_F/A_S$ across species. The regressions across species by PFT were broad and overlapping, with a more general relationship extending across the entire LMA axis. A larger fraction of the variation in $M_F/A_S$ was explained by species-level differences in $A_F/A_S$ (Fig. 3b). As previously noted, $A_F/A_S$ does not differ systematically among PFTs (Fig. 2e), but does vary close to two orders of magnitude across species. In Fig. 3b, the separation among PFTs in $M_F/A_S$ arises due to differences in LMA.

We found that decomposing LMF and LAR as shown in Eq. \ref{eq:M1} was very useful to study climate effects because the second term ($A_S/M_T$) absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB}, Table S1 and Fig. S4) and is otherwise fairly conserved across PFTs (Fig. S4). As shown in Fig. 2b and Fig. 2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and Table S1). $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.

Regardless of how we analyzed the data, we found only weak climatic effects on biomass partitioning within PFTs (Fig. 4). Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table \ref{table/varpartB}). Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) (Fig. 4b-c). These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table S1).

@@

We found that PFTs differed in LMF, and that LMF was proportional to LMA. This conclusion was particularly strong across the three PFTs studied, but it also held across species within PFT. The implication is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) do not differ between PFTs. This result seems robust, as our data include individual plants spanning the entire size range of woody plants in natural forests (0.1-100m) (Fig. 2a). Previous studies have demonstrated a large difference in LMF between angiosperms and gymnosperms \citep{poorter2012}, but have not further separated functional types into deciduous vs. evergreen angiosperms, or provided an explanation in terms of leaf traits. We show that LMA -- a central trait of the leaf economics spectrum (LES) \citep{wrightetal2004} -- explains these differences in a consistent manner.

A previous compilation of plant biomass partitioning \citep{poorter2012} (which included many non field-grown plants) did not find a correlation between LMF and LMA across 8 PFTs (including non-woody PFTs). One possible reason for this discrepancy, apart from differences in sample size and growing conditions, is that they did not correct for plant size in their comparison between LMF and LMA. We found that, as expected, plant size strongly influences LMF (Fig. 2) and LAR (Fig. S3) (see also \citet{poorter2012}). It is thus necessary to correct for plant size when comparing biomass partitioning parameters, as has been noted many times \citep{mcconnaughay1999}. We used a semi-parametric approach to account for plant size, which has the advantage that it does not require an \emph{a priori} assumption on the functional relationship. This was useful because both LMF and LAR showed very non-linear patterns with plant height, even on a logarithmic scale.

@@

% The first sentence here feels a bit repetitive 
We found that, as for LMF, $M_F/A_S$ was also proportional to LMA across PFTs (Fig. 2). Because LMF may be decomposed into the terms $M_F/A_S$ and $M_T/A_S$ (\ref{eq:M2}), this result suggests that $M_T/A_S$ does not vary between PFTs (see also Fig. S4). If this result holds, it means that aboveground biomass may be estimated across PFTs with an equation that does not differ between PFTs. For tropical forests, \citet{chave2005} showed that aboveground biomass can be well estimated as,

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure1.pdf}
    \caption{Overview of the data. (a) Variables were measured on \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species. (b) Coverage of the dataset across global climate space. Grey hexagons indicate the number of 0.5\degree~cells with woody vegetation across the space. Colour symbols show the locations of sampled individuals for three dominant woody functional types.}
    \label{fig:figure1}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
    \caption{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$). Each point is an individual plant. Sample sizes are listed in Table~\ref{table:counttable}}
    \label{fig:figureS2}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
    \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf mass per area. (a) Leaf mass fraction ($M_F/M_T$ = leaf mass / above-ground biomass) by PFT. Each symbol is an individual plant. Lines are generalised additive model fits. (b) and (c) Leaf mass fraction and leaf area ratio ($A_F/M_T$) at the average plant height in the dataset, estimated from panel (a). (d) Average leaf mass per unit basal stem area, and (e) leaf area per unit basal stem area for the three major PFTs confirm that the between-PFT variation in leaf mass fraction is due to leaf mass per unit basal area. Error bars are 95\% confidence intervals. Letters denote significant differences (at $\alpha = 0.05$).}
    \label{fig:figure2}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure4.pdf}
     \caption{Leaf and stem traits capture variation in biomass partitioning across species. Individual plant data were averaged by species and study combinations. Lines are linear regressions with 95\% confidence intervals. Both regressions were significant ($P \textless 0.01$). $R^2$ for fitted relationships are \Sexpr{round(100*r2panela,1)}\% in panel (a) and \Sexpr{round(100*r2panelb,1)}\% in panel (b) (for a linear regression model including PFT as a factor variable).}
  \label{fig:figure4}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
    \caption{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning. a) Leaf mass per unit stem cross-sectional area ($M_F/A_S$) for each PFT within three different biomes.  DA: deciduous angiosperms, EA: evergreen angiosperms, EG: evergreen gymnosperms. Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect. Letters denote significant differences (at $\alpha = 0.05$). b-c) Relationships between climate and $M_F/A_S$ within each PFT. Solid lines are generalised additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
    \label{fig:figure3}
\end{figure}

@@

\textbf{Table S1} Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.\\
\textbf{Table S2} Number of observations used in the analysis of $M_F/A_S$ and $A_F/A_S$.\\
\textbf{Table S3} Number of observations used in the analysis of $A_F/A_S$.\\
\textbf{Table S4} Number of observations used in the analysis of leaf mass fraction ($M_F/M_T$) and leaf area ratio ($A_F/M_T$).\\
\textbf{Table S5} Number of observations used in the analysis of leaf area ratio (LAR).\\
\textbf{Fig. S1} Global coverage of the climate space by the dataset, labelled by vegetation type.\\
\textbf{Fig. S2} Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$).\\
\textbf{Fig. S3} Leaf area ratio ($A_F/M_T$) by PFT.\\
\textbf{Fig. S4} Relationship between above-ground biomass and basal stem area.\\
\textbf{Fig. S5} Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.\\
\textbf{Notes S1} Modelling of biomass partitioning in global vegetation models (GVMs)

>>>>

Here, we use a unique, new database \citep{falster2015} (Fig.~\ref{fig:figoverview}) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms. A second objective was to study relationships between biomass partitioning and climate variables or biome at a global scale. The dataset includes observations of biomass and size metrics for individual plants, compiled from 175 studies across nine vegetation types (Fig. S1), across the three major biomes (boreal, temperate and tropical). In this paper we focus on field-grown woody plants, spanning the entire size range of woody plants (0.01 - \textgreater 100 m height). One key challenge is thus to account for the very large size variation commonly found in any allometric variable \citep{niklas1994}. We do this by fitting a statistical model that does no assume a functional dependence on plant size, allowing us to study PFT patterns at a common plant height.

@@

The raw data in Fig.~\ref{fig:figrawdata} show a steeper increase of woody aboveground biomass ($M_S$) with plant height than for foliage biomass ($M_F$). As a result, there was a strong effect of plant height on LMF , with the three plant functional types clearly differing in LMF across nearly the entire size range (Fig.~\ref{fig:figlmf}a). When corrected for plant height by estimating LMF at a common plant height, we found that LMF was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig.~\ref{fig:figlmf}b). As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig. ~\ref{fig:figlmf}c and Fig. S2).

Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig. ~\ref{fig:figlmf}d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig.~\ref{fig:figlmf}e).

@@

We found that LMF differs predictably among PFTs, however there was considerable variation in all studied variables between species within PFTs (Fig. S4). It is thus pertinent to understand whether we can improve on a PFT-based classification by including other traits that affect biomass partitioning. To do this, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq. \ref{eq:M2}).

@@

The ratio $M_F/A_S$ is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}, using records of stem diameter commonly recorded on long-term monitoring plots. The correlation between $M_F/A_S$ and LMA was found to hold also within PFTs (Fig.~\ref{fig:figlmfspecies}a), explaining 30\% of the variation in $M_F/A_S$ across species. The regressions across species by PFT were broad and overlapping, with a more general relationship extending across the entire LMA axis. A larger fraction of the variation in $M_F/A_S$ was explained by species-level differences in $A_F/A_S$ (Fig.~\ref{fig:figlmfspecies}b). As previously noted, $A_F/A_S$ does not differ systematically among PFTs (Fig.~\ref{fig:figlmf}e), but does vary close to two orders of magnitude across species. In Fig.~\ref{fig:figlmfspecies}b, the separation among PFTs in $M_F/A_S$ arises due to differences in LMA.

We found that decomposing LMF and LAR as shown in Eq. \ref{eq:M1} was very useful to study climate effects because the second term ($A_S/M_T$) absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB}, Table S1 and Fig. S3) and is otherwise fairly conserved across PFTs (Fig. S3). As shown in Fig.~\ref{fig:figlmfspecies}b and Fig.~\ref{fig:figlmf}d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and Table S1). $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.

Regardless of how we analyzed the data, we found only weak climatic effects on biomass partitioning within PFTs (Fig.~\ref{fig:figclimate}). Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table \ref{table/varpartB}). Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) (Fig.~\ref{fig:figclimate}b-c). These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table S1).

@@

We found that PFTs differed in LMF, and that LMF was proportional to LMA. This conclusion was particularly strong across the three PFTs studied, but it also held across species within PFT. The implication is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) do not differ between PFTs. This result seems robust, as our data include individual plants spanning the entire size range of woody plants in natural forests (0.1-100m) (Fig.~\ref{fig:figlmf}a). Previous studies have demonstrated a large difference in LMF between angiosperms and gymnosperms \citep{poorter2012}, but have not further separated functional types into deciduous vs. evergreen angiosperms, or provided an explanation in terms of leaf traits. We show that LMA -- a central trait of the leaf economics spectrum (LES) \citep{wrightetal2004} -- explains these differences in a consistent manner.

A previous compilation of plant biomass partitioning \citep{poorter2012} (which included many non field-grown plants) did not find a correlation between LMF and LMA across 8 PFTs (including non-woody PFTs). One possible reason for this discrepancy, apart from differences in sample size and growing conditions, is that they did not correct for plant size in their comparison between LMF and LMA. We found that, as expected, plant size strongly influences LMF (Fig.~\ref{fig:figlmf}) and LAR (Fig. S3) (see also \citet{poorter2012}). It is thus necessary to correct for plant size when comparing biomass partitioning parameters, as has been noted many times \citep{mcconnaughay1999}. We used a semi-parametric approach to account for plant size, which has the advantage that it does not require an \emph{a priori} assumption on the functional relationship. This was useful because both LMF and LAR showed very non-linear patterns with plant height, even on a logarithmic scale.

@@

% The first sentence here feels a bit repetitive 
We found that, as for LMF, $M_F/A_S$ was also proportional to LMA across PFTs (Fig.~\ref{fig:figlmf}). Because LMF may be decomposed into the terms $M_F/A_S$ and $M_T/A_S$ (\ref{eq:M2}), this result suggests that $M_T/A_S$ does not vary between PFTs (see also Fig. S4). If this result holds, it means that aboveground biomass may be estimated across PFTs with an equation that does not differ between PFTs. For tropical forests, \citet{chave2005} showed that aboveground biomass can be well estimated as,

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure1.pdf}
    \caption{Overview of the data. (a) Variables were measured on \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species. (b) Coverage of the dataset across global climate space. Grey hexagons indicate the number of 0.5\degree~cells with woody vegetation across the space. Colour symbols show the locations of sampled individuals for three dominant woody functional types.}
    \label{fig:figoverview}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
    \caption{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$). Each point is an individual plant. Sample sizes are listed in Table~\ref{table:counttable}}
    \label{fig:figrawdata}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
    \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf mass per area. (a) Leaf mass fraction ($M_F/M_T$ = leaf mass / above-ground biomass) by PFT. Each symbol is an individual plant. Lines are generalised additive model fits. (b) and (c) Leaf mass fraction and leaf area ratio ($A_F/M_T$) at the average plant height in the dataset, estimated from panel (a). (d) Average leaf mass per unit basal stem area, and (e) leaf area per unit basal stem area for the three major PFTs confirm that the between-PFT variation in leaf mass fraction is due to leaf mass per unit basal area. Error bars are 95\% confidence intervals. Letters denote significant differences (at $\alpha = 0.05$).}
    \label{fig:figlmf}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure4.pdf}
     \caption{Leaf and stem traits capture variation in biomass partitioning across species. Individual plant data were averaged by species and study combinations. Lines are linear regressions with 95\% confidence intervals. Both regressions were significant ($P \textless 0.01$). $R^2$ for fitted relationships are \Sexpr{round(100*r2panela,1)}\% in panel (a) and \Sexpr{round(100*r2panelb,1)}\% in panel (b) (for a linear regression model including PFT as a factor variable).}
  \label{fig:figlmfspecies}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
    \caption{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning. a) Leaf mass per unit stem cross-sectional area ($M_F/A_S$) for each PFT within three different biomes.  DA: deciduous angiosperms, EA: evergreen angiosperms, EG: evergreen gymnosperms. Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect. Letters denote significant differences (at $\alpha = 0.05$). b-c) Relationships between climate and $M_F/A_S$ within each PFT. Solid lines are generalised additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
    \label{fig:figclimate}
\end{figure}

@@

\textbf{Table S1} Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.\\
\textbf{Table S5} Number of observations used in the analysis of leaf area ratio (LAR).\\
\textbf{Fig. S1} Global coverage of the climate space by the dataset, labelled by vegetation type.\\
\textbf{Fig. S2} Leaf area ratio ($A_F/M_T$) by PFT.\\
\textbf{Fig. S3} Relationship between above-ground biomass and basal stem area.\\
\textbf{Fig. S4} Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.\\
\textbf{Notes S1} Modelling of biomass partitioning in global vegetation models (GVMs)
